---
title: "STARprotocol"
author: "MAguirrebengoa"
date: "30/08/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
```

## I. Inport data

Read counts from HTseq output for each sample, merge samples into one single dataframe, rename and filter database

```{r}
#1)
HTSC555 = read.table( paste0( path, "Data/Aligned.sortedByCoord555-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC556 = read.table( paste0( path, "Data/Aligned.sortedByCoord556-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC557 = read.table( paste0( path, "Data/Aligned.sortedByCoord557-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC558 = read.table( paste0( path, "Data/Aligned.sortedByCoord558-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC559 = read.table( paste0( path, "Data/Aligned.sortedByCoord559-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC560 = read.table( paste0( path, "Data/Aligned.sortedByCoord560-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC561 = read.table( paste0( path, "Data/Aligned.sortedByCoord561-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC562 = read.table( paste0( path, "Data/Aligned.sortedByCoord562-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC563 = read.table( paste0( path, "Data/Aligned.sortedByCoord563-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC564 = read.table( paste0( path, "Data/Aligned.sortedByCoord564-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC565 = read.table( paste0( path, "Data/Aligned.sortedByCoord565-htseq_count.tabular" ), sep = "\t", header = FALSE );
HTSC566 = read.table( paste0( path, "Data/Aligned.sortedByCoord566-htseq_count.tabular" ), sep = "\t", header = FALSE );

#2)
df = HTSC555;
df = merge( df, HTSC556, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:556 ) );
df = merge( df, HTSC557, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:557 ) );
df = merge( df, HTSC558, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:558 ) );
df = merge( df, HTSC559, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:559 ) );
df = merge( df, HTSC560, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:560 ) );
df = merge( df, HTSC561, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:561 ) );
df = merge( df, HTSC562, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:562 ) );
df = merge( df, HTSC563, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:563 ) );
df = merge( df, HTSC564, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:564 ) );
df = merge( df, HTSC565, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:565 ) );
df = merge( df, HTSC566, by.x = "V1", by.y = "V1", all = TRUE );
colnames( df ) = c( "V1", paste0( "C", 555:566 ) );
rownames( df ) = df[, 1];
df = df[, -1];
dim( df );

#3)
cdf = df[6:nrow( df ), ];
colnames( cdf ) = c( "GFP1", "TOM1", "GFPTOM1", "GFP2", "TOM2", "GFPTOM2", "GFP3", "TOM3", "GFPTOM3", "GFP5", "TOM5", "GFPTOM5" );

tableCleanNo0 = cdf[rowSums( cdf )  >= 55, ]
dim( cdf );
dim( tableCleanNo0 );
```


## II. Explore raw data

### 1. Control Dataframe content with summary
```{r}
summary( cdf );
summary( tableCleanNo0 );
```

### 2. Proportion of null counts in the table not cleaned
```{r}
apply(cdf, 2, function(x){return( sum( x == 0) * 100 / length( x ) ) } );
apply(tableCleanNo0, 2, function(x){return( sum( x == 0) * 100 / length( x ) ) } );

```

### 4.3. Density Plot
```{r}
par( mfrow = c( 3, 4 ) )
for( c in 1:ncol( cdf ) ){
  hist( log2( cdf[, c] + 1 ), main = colnames(cdf)[c] )
}
for( c in 1:ncol( cdf ) ){
  hist( log2( tableCleanNo0[, c] + 1 ), main = colnames(cdf)[c] )
}
```

### 4.4. Most expressed gene
```{r}
rbind( colnames(cdf ), rownames( tableCleanNo0 )[apply(tableCleanNo0, 2, function(x){return( which( x == max( x ) ) )})] )
```

### 4.5. Pairwise scatter plot
```{r}
# pairs( tableCleanNo0 )
pairs( tableCleanNo0[, paste0( "GFP", c( 1:3, 5 ) )] )
pairs( tableCleanNo0[, paste0( "TOM", c( 1:3, 5 ) )] )
pairs( tableCleanNo0[, paste0( "GFPTOM", c( 1:3, 5 ) )] )

```

### 4.6. CAH
On raw data (Euclidian-Ward.D)
```{r}
plot( hclust( dist( t( cdf ), method = "euclidean" ),  method = "ward.D" ), hang = -1 );
```

On clean data (Euclidian-Ward.D)
```{r}
plot( hclust( dist( t( tableCleanNo0 ), method = "euclidean" ),  method = "ward.D" ), hang = -1 );
```

On normalized data see below...

### 4.7. PCA
On raw data
```{r}
#BiocManager::install("mixOmics")
library( mixOmics )
pca = pca( t( cdf ) );
plotIndiv( pca )
```
On clean data
```{r}
pca = pca( t( tableCleanNo0 ) );
plotIndiv( pca, pch= 16, cex= 1 )
```
On clean data scaled
```{r}
pca = pca( t( tableCleanNo0  ), center = TRUE, scale = TRUE );
plotIndiv( pca )
```


### 5.1. Build design and run normalization

```{r}
#BiocManager::install("DESeq2")
library(DESeq2)
fdf = cdf[rowSums( cdf ) >= 55, ];
dim( fdf ); # [1] 37274     12
colnames( fdf ) = c( "GFP1", "TOM1", "GFPTOM1", "GFP2", "TOM2", "GFPTOM2", "GFP3", "TOM3", "GFPTOM3", "GFP5", "TOM5", "GFPTOM5" );

#2)
colData = data.frame( "condition" = rep( c( "GFP", "Tom", "GFPTom" ), 4 ), "replicat" = rep( paste0( "REP", c( 1:3, 5 ) ), each = 3 ) );
rownames( colData ) = colnames( fdf );
ddsBF = DESeqDataSetFromMatrix( countData = fdf, colData = colData, design = ~condition+replicat );
ds = DESeq( ddsBF );

#3)
res_IFF_BF_GFPTOMvsGFP = results( dsBF, independentFiltering = FALSE, cooksCutoff = FALSE, contrast = c( "condition", "GFPTom", "GFP" ) );
res_IFF_BF_GFPTOMvsTOM = results( dsBF, independentFiltering = FALSE, cooksCutoff = FALSE, contrast = c( "condition", "GFPTom", "Tom" ) );
res_IFF_BF_TOMvsGFP = results( dsBF, independentFiltering = FALSE, cooksCutoff = FALSE, contrast = c( "condition", "Tom", "GFP" ) );

#4)
m1 = merge( fdf,  counts(dsBF, normalized = TRUE ), by = "row.names", all = TRUE );
m2 = merge( data.frame( res_IFF_BF_GFPTOMvsGFP ), data.frame( res_IFF_BF_GFPTOMvsTOM ), by = "row.names", all = TRUE );
rownames( m1 ) = m1[, "Row.names"];
m1 = m1[, -1];
colnames( m1 ) = paste( c( rep( "Raw", 12 ), rep( "Norm", 12 ) ), rep( colnames( fdf ), 2 ), sep = "_" );
rownames( m2 ) = m2[, "Row.names"];
m2 = m2[, -1];
m3 = merge( m2, data.frame( res_IFF_BF_TOMvsGFP ), by = "row.names", all = TRUE );
rownames( m3 ) = m3[, 1]; 
m3 = m3[, -1];
colnames( m3 ) = paste( rep( c( "DGE1_GFPTomvsGFP", "DGE2_GFPTomvsTom", "DGE3_TomvsGFP" ), each = 6 ), rep( c( "baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj" ), 3 ), sep = "_" );
finatab = merge( m1, m3, by = "row.names", all = TRUE );
rownames( finatab ) = finatab[, "Row.names"];
finatab = finatab[, -1];
```

### 5.2. Control Normalization

#### 5.2.1. Normalisation
```{r}
sizeFactors( ds )
boxplot( log2( tableCleanNo0 + 1 ) )
boxplot( log2( counts( ds, normalized = TRUE ) + 1 ) )
```

#### 5.2.2. CAH & clustering on normalized data
```{r}
plot( hclust( dist( t( counts( ds, normalized = TRUE ) ), method = "euclidean" ),  method = "ward.D" ), hang = -1 );
pca = pca( t( counts( ds, normalized = TRUE ) ) )
plotIndiv( pca, pch= 16, cex= 1)
```

#### 5.2.3. CAH & clustering with VST
```{r}
vst = vst( ddsBF )
plot( hclust( dist( t( assay( vst ) ), method = "euclidean" ),  method = "ward.D" ), hang = -1 );
pca = pca( t( assay( vst ) ) )
plotIndiv( pca)
```


### 5.4. Plots

#### 5.4.1. Basics
MAplot
```{r}
plot( log2( res_IFF_BF_GFPTOMvsGFP$baseMean ), res_IFF_BF_GFPTOMvsGFP$log2FoldChange, xlab = "log2 baseMean", ylab = "log2FC", ylim = c( -10, +10 ) )
points(log2( res_IFF_BF_GFPTOMvsGFP$baseMean[res_IFF_BF_GFPTOMvsGFP$padj < 0.05] ), res_IFF_BF_GFPTOMvsGFP$log2FoldChange[res_IFF_BF_GFPTOMvsGFP$padj < 0.05], pch = 16, col = "red" )

plot( log2( res_IFF_BF_GFPTOMvsTOM$baseMean ), res_IFF_BF_GFPTOMvsTOM$log2FoldChange, xlab = "log2 baseMean", ylab = "log2FC", ylim = c( -10, +10 ) )
points(log2( res_IFF_BF_GFPTOMvsTOM$baseMean[res_IFF_BF_GFPTOMvsTOM$padj < 0.05] ), res_IFF_BF_GFPTOMvsTOM$log2FoldChange[res_IFF_BF_GFPTOMvsTOM$padj < 0.05], pch = 16, col = "red" )

plot( log2( res_IFF_BF_TOMvsGFP$baseMean ), res_IFF_BF_TOMvsGFP$log2FoldChange, xlab = "log2 baseMean", ylab = "log2FC" , ylim = c( -10, +10 ))
points(log2( res_IFF_BF_TOMvsGFP$baseMean[res_IFF_BF_TOMvsGFP$padj < 0.05] ), res_IFF_BF_TOMvsGFP$log2FoldChange[res_IFF_BF_TOMvsGFP$padj < 0.05], pch = 16, cex = 1, col = "red" )
```

Volcanoplot
```{r}
plot( res_IFF_BF_GFPTOMvsGFP$log2FoldChange, -1*log10( res_IFF_BF_GFPTOMvsGFP$padj ), xlab = "log2 baseMean", ylab = "log2FC" )
points( res_IFF_BF_GFPTOMvsGFP$log2FoldChange[res_IFF_BF_GFPTOMvsGFP$padj < 0.05], -1*log10( res_IFF_BF_GFPTOMvsGFP$padj )[res_IFF_BF_GFPTOMvsGFP$padj < 0.05], pch = 16, col = "red" )

plot( res_IFF_BF_GFPTOMvsTOM$log2FoldChange, -1*log10( res_IFF_BF_GFPTOMvsTOM$padj ), xlab = "log2 baseMean", ylab = "log2FC" )
points( res_IFF_BF_GFPTOMvsTOM$log2FoldChange[res_IFF_BF_GFPTOMvsTOM$padj < 0.05], -1*log10( res_IFF_BF_GFPTOMvsTOM$padj )[res_IFF_BF_GFPTOMvsTOM$padj < 0.05], pch = 16, col = "red" )

plot( res_IFF_BF_TOMvsGFP$log2FoldChange, -1*log10( res_IFF_BF_TOMvsGFP$padj ), xlab = "log2 baseMean", ylab = "log2FC" )
points( res_IFF_BF_TOMvsGFP$log2FoldChange[res_IFF_BF_TOMvsGFP$padj < 0.05], -1*log10( res_IFF_BF_TOMvsGFP$padj )[res_IFF_BF_TOMvsGFP$padj < 0.05], pch = 16, col = "red" )
```